A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-02-08 paper/paper.Rmd
refining conclusions

Finally, for the data providers we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource. The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs. These values can be validated with simple checks at data entry. Providing an open data resource also is accompanied with the responsibility that the data, especially data that is as valuable as this, is reliable. Users need to be able to trust the data.

Why is this exercise important for teachers and students? The data package and software could be used for illustrating data cleaning in the classroom. The data provided in the package can be used as examples i nthe teaching of initial data analysis, and modeling, and exploratory data analysis of longitudinal data.

XXX Importance of public policy vs for the public; and the importance of using relatable data examples in teaching that it needs to be acurate, and limitations on interpretation need to be available.

>>>>

For the data providers, we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource. The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs. These values can be validated with simple checks at data entry. Providing an open data resource also is accompanied with the responsibility that the data, especially data that is as valuable as this, is reliable. Users need to be able to trust the data.

Why is this exercise important for teachers and students? The data package and software could be used for illustrating data cleaning in the classroom. The refreshed and contemporary data provided in the package can be used as examples in the teaching of initial data analysis and modeling, and exploratory data analysis of longitudinal data.

The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual. From modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns. That more education leads to increasingly higher wages, which is a satisfying result for educators. It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that race matters and that being black leads to lower wages. This is a disturbing finding because there is no rational for such a difference in a fair society. This is would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

These are interpretations of results from analyzing the wages data, which rely on trusting that the data provided is accurate and valid. The data is collected from a reputable organization, but the data has obvious errors that need correcting. This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.

Having trustworthy data is imperative for statistics and data science education. Students do interpret results and take seriously what is reported from example analyses. If the data in the examples are synthetic, with our own inherent prejudices (on fog e.g. gender and race) applied, or inaccurately processed leaving errors, we as educators are being irresponsible because the societal message taught to students is flawed. This paper demonstrates the process to produce a trustworthy data set for teaching.

### 2022-02-08 paper/paper.tex
refining conclusions

Finally, for the data providers we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource. The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs. These values can be validated with simple checks at data entry. Providing an open data resource also is accompanied with the responsibility that the data, especially data that is as valuable as this, is reliable. Users need to be able to trust the data.

Why is this exercise important for teachers and students? The data package and software could be used for illustrating data cleaning in the classroom. The data provided in the package can be used as examples i nthe teaching of initial data analysis, and modeling, and exploratory data analysis of longitudinal data.

XXX Importance of public policy vs for the public; and the importance of using relatable data examples in teaching that it needs to be acurate, and limitations on interpretation need to be available.

>>>>

For the data providers, we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource. The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs. These values can be validated with simple checks at data entry. Providing an open data resource also is accompanied with the responsibility that the data, especially data that is as valuable as this, is reliable. Users need to be able to trust the data.

Why is this exercise important for teachers and students? The data package and software could be used for illustrating data cleaning in the classroom. The refreshed and contemporary data provided in the package can be used as examples in the teaching of initial data analysis and modeling, and exploratory data analysis of longitudinal data.

The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual. From modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns. That more education leads to increasingly higher wages, which is a satisfying result for educators. It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that race matters and that being black leads to lower wages. This is a disturbing finding because there is no rational for such a difference in a fair society. This is would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

These are interpretations of results from analyzing the wages data, which rely on trusting that the data provided is accurate and valid. The data is collected from a reputable organization, but the data has obvious errors that need correcting. This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.

Having trustworthy data is imperative for statistics and data science education. Students do interpret results and take seriously what is reported from example analyses. If the data in the examples are synthetic, with our own inherent prejudices (on fog e.g.~gender and race) applied, or inaccurately processed leaving errors, we as educators are being irresponsible because the societal message taught to students is flawed. This paper demonstrates the process to produce a trustworthy data set for teaching.
